\documentclass[acmsmall,screen,nonacm]{acmart}

\usepackage{macros}
\usepackage{mathpartir}
\usepackage{cleveref}
\usepackage{tikz}
\usepackage{minted}
\setminted{fontsize=\small, frame=none, breaklines}

\begin{document}

\title{Operational Semantics of the \textsf{alifib} Interpreter}
\author{} \date{}

\begin{abstract}
We give a big-step operational semantics for the \textsf{alifib}
interpreter.  The language constructs \emph{regular directed complexes}---combinatorial
structures whose cells and pasting diagrams are modelled by
\emph{molecules}, an inductively defined class of shapes---and computes
with pasting diagrams over them, for the purpose of higher-dimensional
rewriting and higher category theory.  We describe the abstract syntax, the interpreter state,
and give evaluation rules for every language construct.  We draw attention
to several subtle design decisions and their rationale.
\end{abstract}

\maketitle

\tableofcontents

% ===========================================================================
\section{Introduction}
\label{sec:intro}
% ===========================================================================

\textsf{alifib} is a language for constructing \emph{regular directed
complexes}\footnote{A.\ Hadzihasanovic, \emph{Combinatorics of
higher-categorical diagrams}, 2024; D.-M.\ Kessler,
\emph{Computational Aspects of Rewriting in Higher-Dimensional
Diagrams}, 2025.} and computing with \emph{pasting diagrams} over
them, in the framework of higher-dimensional rewriting.

The shapes of pasting diagrams in this framework are called
\emph{molecules}: an inductively defined class built from points by
pasting and the atom construction (\cref{sec:molecules:def}).  A
finite \emph{regular directed complex} is an oriented graded poset
whose cells are ``locally atoms'': the closure of every cell is an
atom.  In full generality, directed complexes are presheaves on the
category of atoms; \textsf{alifib} works with the finite case,
constructing them cell by cell.  Regularity ensures that every pasting
diagram decomposes uniquely into atoms, one per cell, making the
labelling of diagrams well-defined.

A \emph{pasting diagram} over a regular directed complex~$P$ is a
composable configuration of cells drawn from~$P$, whose shape is a
\emph{molecule}.  Users build pasting diagrams by composing cells;
these diagrams serve as the source and target boundaries of new,
higher-dimensional cells.  \Cref{sec:molecules} gives formal
definitions and illustrative examples, including concrete
\textsf{alifib} code.

Beyond constructing directed complexes, \textsf{alifib} manages
\emph{partial maps} (structure-preserving maps between complexes) and
\emph{modules} (file-level namespaces).  The interpreter maintains a
mutable global state and threads it through the evaluation of each
instruction.

This document provides a self-contained formal account of the
interpreter's behaviour, abstracting away Rust-level implementation
details (error reporting, span tracking, \texttt{Arc}-based sharing)
while preserving every semantically meaningful step.

\paragraph{Conventions.}
Semantic domains and metavariable conventions are collected in
\cref{fig:domains}.  We use $\Downarrow$ for big-step evaluation.
Metavariables are colour-coded: \Type{types} in blue,
\Cell{cells} in brown, \Map{maps} in teal, \Diag{diagrams} in
purple, \St{state components} in orange, and \Id{identifiers/tags} in
green.  Keywords appear in $\kw{sans\text{-}serif}$.
All nonterminals in the grammar are set in \textsf{sans-serif}.
For partial maps and lookups, $f(x)\dn$ means $f$ is defined at~$x$
and $f(x)\up$ means it is undefined.

% ===========================================================================
\section{Directed Complexes and Diagrams}
\label{sec:molecules}
% ===========================================================================

This section recalls the combinatorial framework underlying
\textsf{alifib}: molecules, regular directed complexes, and diagrams.
We describe the abstract objects only; their implementation-level
representation is given in \cref{sec:data}.

The basic objects are \emph{cells}, which have a dimension $k \geq 0$.
Each $k$-cell with $k > 0$ has an \emph{input boundary} and an
\emph{output boundary}, both consisting of $(k{-}1)$-cells.  Input
faces are ``source'' faces and output faces are ``target'' faces.

% ---------------------------------------------------------------------------
\subsection{Molecules}
\label{sec:molecules:def}
% ---------------------------------------------------------------------------

\emph{Molecules} are the shapes of pasting diagrams.  They form an
inductively defined class of oriented graded posets:
\begin{enumerate}
\item \textbf{Point.}  The ogposet with a single 0-cell is a molecule.
\item \textbf{Paste.}  If $U$ and $V$ are molecules and
  $\bdy{+}{k}{U} \cong \bdy{-}{k}{V}$ for some $k \geq 0$, then the
  pushout $U \paste{k} V$ (gluing $U$ and $V$ along the shared
  $k$-boundary) is a molecule.
\item \textbf{Atom.}  If $U$ and $V$ are \emph{round} molecules of
  the same dimension with matching boundaries, then $U \Rightarrow V$
  (the ogposet formed by adding a single top-dimensional cell with
  input boundary~$U$ and output boundary~$V$) is a molecule.  The
  result is called an \emph{atom}.
\end{enumerate}

\begin{figure}[t]
\centering
\begin{tikzpicture}[
  cell/.style={fill, circle, inner sep=1.5pt},
  >=stealth,
  every node/.style={font=\small}
]
% (a) Point
\begin{scope}[shift={(0,0)}]
  \node[cell, label=below:{$a$}] at (0,0) {};
  \node[anchor=north] at (0,-0.5) {\footnotesize (a) point};
\end{scope}

% (b) Arrow (1-globe)
\begin{scope}[shift={(2,0)}]
  \node[cell] (a) at (0,0) {};
  \node[cell] (b) at (1.5,0) {};
  \draw[->] (a) -- node[above] {$f$} (b);
  \node[anchor=north] at (0.75,-0.5) {\footnotesize (b) atom: 1-globe};
\end{scope}

% (c) 2-globe
\begin{scope}[shift={(5.5,0)}]
  \node[cell] (a) at (0,0) {};
  \node[cell] (b) at (2,0) {};
  \draw[->, bend left=25] (a) to node[above] {$f$} (b);
  \draw[->, bend right=25] (a) to node[below] {$g$} (b);
  \node at (1,0) {$\Downarrow\!\alpha$};
  \node[anchor=north] at (1,-0.7) {\footnotesize (c) atom: 2-globe};
\end{scope}

% (d) f #_0 g
\begin{scope}[shift={(0.5,-2.5)}]
  \node[cell] (a) at (0,0) {};
  \node[cell] (b) at (1.5,0) {};
  \node[cell] (c) at (3,0) {};
  \draw[->] (a) -- node[above] {$f$} (b);
  \draw[->] (b) -- node[above] {$g$} (c);
  \node[anchor=north] at (1.5,-0.5) {\footnotesize (d) paste: $f \paste{0} g$};
\end{scope}

% (e) alpha #_1 beta
\begin{scope}[shift={(5.5,-2.5)}]
  \node[cell] (a) at (0,0) {};
  \node[cell] (b) at (2,0) {};
  \draw[->, bend left=40] (a) to node[above] {$f$} (b);
  \draw[->] (a) -- node[fill=white, inner sep=1pt] {$g$} (b);
  \draw[->, bend right=40] (a) to node[below] {$h$} (b);
  \node at (0.7,0.35) {\scriptsize$\Downarrow\!\alpha$};
  \node at (0.7,-0.35) {\scriptsize$\Downarrow\!\beta$};
  \node[anchor=north] at (1,-1) {\footnotesize (e) paste: $\alpha \paste{1} \beta$};
\end{scope}
\end{tikzpicture}
\caption{Basic molecules.  (a)~A point.  (b)~The 1-globe
  (atom of two points).  (c)~The 2-globe (atom of two 1-globes).
  (d)~A composable pair of arrows (paste at dimension~0).
  (e)~Vertical composition of two 2-cells (paste at dimension~1).}
\label{fig:molecules:basic}
\end{figure}

\Cref{fig:molecules:basic} illustrates these constructors.  The
point~(a) is the base case.  The arrow~(b) is the atom
$\{a\} \Rightarrow \{b\}$ formed from two points; the 2-globe~(c) is
the atom $f \Rightarrow g$ formed from two arrows with matching
endpoints.  Shape~(d) is a paste at dimension~0: two arrows $f$ and $g$
glued along the shared point where $f$ ends and $g$ begins.  Shape~(e)
is a paste at dimension~1: two 2-cells $\alpha$ and $\beta$ sharing the
intermediate arrow~$g$.

% ---------------------------------------------------------------------------
\subsection{Properties: Pure, Round, and Globular}
\label{sec:molecules:props}
% ---------------------------------------------------------------------------

Not all molecules are equally well-behaved.  The key hierarchy is:
\[
  \text{globe} \;\subset\; \text{round molecule}
  \;\subset\; \text{pure molecule}
  \;\subset\; \text{molecule}
\]

\paragraph{Globe (atom).}
A molecule is a \emph{globe} if it has a unique greatest
element---equivalently, if it is produced by the point or atom
constructors alone, without pasting.  The $n$-globe is the shape of a
single $n$-cell with its full boundary.

\paragraph{Round molecule.}
A molecule is \emph{round} if, at every dimension $k$ below the top,
the interiors of the input and output $k$-boundaries are disjoint:
no cell appears as both a ``source'' and a ``target'' at the same
dimension.  All globes are round; so is any vertical paste
$U \paste{k} V$ where $k = \dim(U) - 1$.  Roundness matters because
the atom constructor \emph{requires} its source and target to be round,
ensuring that the resulting cell has well-defined boundaries.

\paragraph{Pure molecule.}
A molecule is \emph{pure} if all its maximal cells (those with no
cofaces) have the same dimension.  For finite-dimensional molecules,
roundness implies purity, but not conversely.

\begin{figure}[t]
\centering
\begin{tikzpicture}[
  cell/.style={fill, circle, inner sep=1.5pt},
  >=stealth,
  every node/.style={font=\small}
]
% (a) Globe
\begin{scope}[shift={(0,0)}]
  \node[cell, label=left:{\scriptsize$a$}] (a) at (0,0) {};
  \node[cell, label=right:{\scriptsize$b$}] (b) at (1.8,0) {};
  \draw[->, bend left=25] (a) to node[above] {$f$} (b);
  \draw[->, bend right=25] (a) to node[below] {$g$} (b);
  \node at (0.9,0) {$\Downarrow\!\alpha$};
  \node[anchor=north] at (0.9,-0.7) {\footnotesize (a) globe};
\end{scope}

% (b) Round, not globe
\begin{scope}[shift={(4,0)}]
  \node[cell, label=left:{\scriptsize$a$}] (a) at (0,0) {};
  \node[cell, label=right:{\scriptsize$b$}] (b) at (1.8,0) {};
  \draw[->, bend left=40] (a) to node[above] {$f$} (b);
  \draw[->] (a) -- node[fill=white, inner sep=1pt] {$g$} (b);
  \draw[->, bend right=40] (a) to node[below] {$h$} (b);
  \node at (0.65,0.35) {\scriptsize$\alpha$};
  \node at (0.65,-0.35) {\scriptsize$\beta$};
  \node[anchor=north] at (0.9,-0.85) {\footnotesize (b) round, not globe};
\end{scope}

% (c) Pure, not round
\begin{scope}[shift={(8,0)}]
  \node[cell, label=left:{\scriptsize$a$}] (a) at (0,0) {};
  \node[cell, label=above:{\scriptsize$b$}] (b) at (1.5,0) {};
  \node[cell, label=right:{\scriptsize$c$}] (c) at (3,0) {};
  \draw[->, bend left=20] (a) to node[above] {$f$} (b);
  \draw[->, bend right=20] (a) to node[below] {$g$} (b);
  \draw[->, bend left=20] (b) to node[above] {$h$} (c);
  \draw[->, bend right=20] (b) to node[below] {$k$} (c);
  \node at (0.6,0) {\scriptsize$\alpha$};
  \node at (2.4,0) {\scriptsize$\beta$};
  \node[anchor=north] at (1.5,-0.7) {\footnotesize (c) pure, not round};
\end{scope}

% (d) Not pure
\begin{scope}[shift={(4,-3)}]
  \node[cell, label=left:{\scriptsize$a$}] (a) at (0,0) {};
  \node[cell, label=above:{\scriptsize$b$}] (b) at (1.5,0) {};
  \node[cell, label=right:{\scriptsize$c$}] (c) at (3,0) {};
  \draw[->, bend left=25] (a) to node[above] {$f$} (b);
  \draw[->, bend right=25] (a) to node[below] {$g$} (b);
  \draw[->] (b) -- node[above] {$h$} (c);
  \node at (0.6,0) {\scriptsize$\alpha$};
  \node[anchor=north] at (1.5,-0.6) {\footnotesize (d) not pure};
\end{scope}
\end{tikzpicture}
\caption{The molecule hierarchy.  (a)~The 2-globe is a globe (one
  top-dimensional cell).  (b)~Vertical composition $\alpha \paste{1}
  \beta$ is round but not a globe (two top-dimensional cells).
  (c)~Horizontal composition $\alpha \paste{0} \beta$ is pure (all
  maximal cells are 2-dimensional) but not round: the point~$b$ lies
  in the interior of both the input and output 1-boundaries.
  (d)~The paste $\alpha \paste{0} h$ is not pure: the maximal cells
  $\alpha$ and $h$ have dimensions~2 and~1.}
\label{fig:molecules:hierarchy}
\end{figure}

\Cref{fig:molecules:hierarchy} illustrates the distinctions.  In~(c),
the horizontal composite $\alpha \paste{0} \beta$ is pure (both maximal
cells have dimension~2) but fails roundness: the interior point~$b$ is a
face of both input-extremal 1-cells ($f$ and $h$) and output-extremal
1-cells ($g$ and $k$), so it belongs to the interior of both the input
and output 1-boundaries.  In~(d), the paste $\alpha \paste{0} h$ is not
pure because $\alpha$ (dimension~2) and $h$ (dimension~1) are both
maximal.

% ---------------------------------------------------------------------------
\subsection{Regular Directed Complexes}
\label{sec:molecules:regular}
% ---------------------------------------------------------------------------

Molecules are ``locally atoms'': for any element~$x$ of a molecule~$U$,
the closure $\mathrm{cl}\{x\}$ is an atom.  Turning this local property
into a definition gives the class of regular directed complexes.

A \emph{regular directed complex} is an oriented graded poset~$P$ such
that, for all $x \in P$, $\mathrm{cl}\{x\}$ is an atom.  Given such
a~$P$, the set $\mathit{Mol}/P$ of isomorphism classes of morphisms
$f \colon U \to P$, with~$U$ ranging over molecules, admits the
structure of a strict $\omega$-category whose generators are in
bijection with the elements of~$P$.  A \emph{diagram of shape~$U$
in~$P$} is such a morphism; it is a \emph{pasting diagram} when $U$ is
a molecule, and a \emph{cell} when $U$ is an atom.

In full generality, a \emph{directed complex} is a presheaf on the
category of atoms (equivalently, a $\Gamma$-continuous presheaf on
finite regular directed complexes and local embeddings).  Every finite
regular directed complex is representable in this category.
\textsf{alifib} works with the finite case, constructing regular
directed complexes cell by cell.

\textsf{alifib} enforces regularity by construction.  Every cell
declaration falls into one of two cases:
\begin{itemize}
  \item A \emph{0-cell} (bare name): $\mathrm{cl}\{x\}$ is a point,
    which is an atom.
  \item A \emph{higher cell} ($n : d^- \to d^+$): the interpreter
    checks that $d^-$ and $d^+$ are round, then constructs
    $\mathrm{cl}\{n\}$ as the atom $d^- \Rightarrow d^+$.
\end{itemize}
Since every cell's closure is an atom, the resulting oriented graded
poset is always a regular directed complex.

% ---------------------------------------------------------------------------
\subsection{Examples in \textsf{alifib}}
\label{sec:molecules:examples}
% ---------------------------------------------------------------------------

We illustrate how \textsf{alifib} declarations build regular directed
complexes.  These examples use the concrete syntax informally; the
formal grammar is given in \cref{sec:syntax}.

\paragraph{A 2-globe.}
The simplest non-trivial example declares a 2-cell between two
parallel arrows:
\begin{minted}{alifib}
@Type
TwoGlobe <<= {
  a, b,
  f: a -> b,
  g: a -> b,
  alpha: f -> g
}
\end{minted}
This produces the regular directed complex in
\cref{fig:example:complexes}(a), with five cells: two 0-cells $a, b$,
two 1-cells $f, g : a \to b$, and one 2-cell
$\alpha : f \Rightarrow g$.  Each declaration corresponds to an atom
constructor: \texttt{a} and~\texttt{b} are points;
\texttt{f} and~\texttt{g} are 1-globes $\{a\} \Rightarrow \{b\}$;
and \texttt{alpha} is the 2-globe $f \Rightarrow g$.

\paragraph{A triangle.}
A more interesting example declares a 2-cell whose source boundary is
a composite:
\begin{minted}{alifib}
@Type
Triangle <<= {
  a, b, c,
  f: a -> b,
  g: b -> c,
  h: a -> c,
  alpha: f g -> h
}
\end{minted}
Here \texttt{f g} is the implicit paste $f \paste{0} g$ (two arrows
sharing the point~$b$).  The resulting directed complex
(\cref{fig:example:complexes}(b)) has seven cells.  The atom
$\mathrm{cl}\{\alpha\}$ has the non-globular shape
$(f \paste{0} g) \Rightarrow h$: its input boundary is a composite,
not a single cell.  Roundness of $f \paste{0} g$ and $h$ is checked
automatically.

\begin{figure}[t]
\centering
\begin{tikzpicture}[
  cell/.style={fill, circle, inner sep=1.5pt},
  >=stealth,
  every node/.style={font=\small}
]
% (a) 2-globe
\begin{scope}[shift={(0,0)}]
  \node[cell, label=left:{$a$}] (a) at (0,0) {};
  \node[cell, label=right:{$b$}] (b) at (2.5,0) {};
  \draw[->, bend left=25] (a) to node[above] {$f$} (b);
  \draw[->, bend right=25] (a) to node[below] {$g$} (b);
  \node at (1.25,0) {$\Downarrow\!\alpha$};
  \node[anchor=north] at (1.25,-0.7) {\footnotesize (a) \texttt{TwoGlobe}};
\end{scope}
% (b) Triangle
\begin{scope}[shift={(6,0)}]
  \node[cell, label=left:{$a$}] (a) at (0,0) {};
  \node[cell, label=right:{$c$}] (c) at (3,0) {};
  \node[cell, label=above:{$b$}] (b) at (1.5,1.2) {};
  \draw[->] (a) to node[above left=-2pt] {$f$} (b);
  \draw[->] (b) to node[above right=-2pt] {$g$} (c);
  \draw[->] (a) to node[below] {$h$} (c);
  \node at (1.5,0.35) {$\Downarrow\!\alpha$};
  \node[anchor=north] at (1.5,-0.7) {\footnotesize (b) \texttt{Triangle}};
\end{scope}
\end{tikzpicture}
\caption{Two regular directed complexes declared in \textsf{alifib}.
  (a)~The 2-globe: five cells, all closures are globes.
  (b)~A triangle: seven cells; $\mathrm{cl}\{\alpha\}$ has the
  non-globular atom shape $(f \paste{0} g) \Rightarrow h$.}
\label{fig:example:complexes}
\end{figure}

\paragraph{Vocabulary.}
A \texttt{@Type} block introduces one or more types. The syntax
$n \;\texttt{<<=}\; \{\ldots\}$ declares a type named~$n$ with a
complex body.  Inside the body, a bare name declares a 0-cell; a
name with boundary $n\texttt{: } d^- \texttt{ -> } d^+$ declares a
higher cell; and juxtaposition of names denotes implicit pasting.

% ===========================================================================
\section{Representation}
\label{sec:data}
% ===========================================================================

The interpreter represents the abstract objects of
\cref{sec:molecules} using concrete data structures.  Molecules (the
shapes of pasting diagrams) become \emph{oriented graded posets}
(ogposets)---finite graded sets with face and coface maps.  Pasting
diagrams (morphisms $U \to X$ with $U$ a molecule) become
\emph{diagrams}---an ogposet shape labelled with tags, together with
paste trees recording how the diagram was assembled.  Maps of directed
complexes become \emph{partial maps}---tag-indexed tables of image
diagrams.  Regular directed complexes themselves become
\emph{complexes}---namespaces of generators, named diagrams, and maps
(\cref{sec:state:complex}).

% ---------------------------------------------------------------------------
\subsection{Oriented Graded Posets}
\label{sec:data:ogposet}
% ---------------------------------------------------------------------------

An \emph{oriented graded poset} (ogposet) $G$ consists of:
\begin{itemize}
  \item A \emph{dimension} $\dim(G) \in \mathbb{Z}_{\geq -1}$
    (where $-1$ indicates the empty ogposet).
  \item For each $0 \leq d \leq \dim(G)$, a finite set of
    \emph{cells} $G_d$.
  \item \emph{Face maps}
    $\bdyin{d}, \bdyout{d} : G_d \to \mathcal{P}(G_{d-1})$
    giving the input and output $(d{-}1)$-cells on the boundary
    of each $d$-cell.
  \item Dually, \emph{coface maps}
    recording which higher-dimensional cells each cell
    appears in the boundary of.
\end{itemize}
An ogposet represents the unlabelled shape of a molecule or, more
generally, of any sub-ogposet of a regular directed complex.  In the
Rust implementation, cells at each dimension are indexed by position
($0, 1, \ldots$) and face/coface sets are stored as sorted vectors of
indices.

\paragraph{Operations.}
\begin{itemize}
  \item \textbf{Boundary extraction.}
    $\bdy{\pm}{k}{G}$ extracts the input ($-$) or output ($+$)
    $k$-boundary of $G$ as a sub-ogposet, together with an
    \emph{embedding} recording which cells of $G$ appear in the
    boundary. The embedding consists of a forward map and an
    inverse map, both indexed by dimension.

  \item \textbf{Normalisation.}
    $\mathsf{norm}(G)$ produces a canonical reordering of the cells
    by traversing from the input-extremal cells (those with no input
    cofaces). Two ogposets are isomorphic if and only if they have
    the same normal form. Both normalisation and boundary extraction
    are cached by pointer identity.

  \item \textbf{Isomorphism.}
    $\mathsf{iso}(G_1, G_2)$ checks whether $G_1 \cong G_2$ by
    normalising both and comparing the canonical forms, then
    composing the two embeddings to produce a direct isomorphism
    $G_1 \to G_2$.

  \item \textbf{Roundness.}
    $G$ is \emph{round} if the input and output interiors are
    disjoint at every dimension---a necessary condition for $G$ to
    serve as the source or target of an atom
    (\cref{sec:molecules:props}).
\end{itemize}

% ---------------------------------------------------------------------------
\subsection{Diagrams}
\label{sec:data:diagram}
% ---------------------------------------------------------------------------

Abstractly, a pasting diagram in a directed complex~$X$ is a morphism
$u \colon U \to X$ where $U$ is a molecule (\cref{sec:molecules:regular}).
The interpreter represents such a diagram as a triple $\Diag{D}$:
\begin{itemize}
  \item A \emph{shape}: an ogposet $G$ (representing the molecule~$U$,
    shared via reference counting).
  \item \emph{Labels}:
    for each dimension $d$, a map
    $\ell_d : G_d \to \Tag$
    assigning a tag to every cell in the shape (encoding the morphism
    $u$ by recording which cell of~$X$ each position maps to).
  \item \emph{Paste trees}: for each dimension $d$ and sign
    $s \in \{-,+\}$, a tree recording how the $d$-level structure
    was built by pasting.
\end{itemize}

The shape represents the combinatorial structure of the molecule; the
labels record \emph{which} generators of the directed complex occupy
each position; the paste trees record \emph{how} the diagram was
assembled.

\paragraph{Paste trees.}
A paste tree is defined inductively:
\[
  \mathit{PT} \;\Coloneqq\;
    \mathsf{Leaf}(t) \mid
    \mathsf{Node}(k, \mathit{PT}_L, \mathit{PT}_R)
\]
A $\mathsf{Leaf}(t)$ records a single generator with tag $t$.
A $\mathsf{Node}(k, L, R)$ records a paste at dimension $k$
of the sub-diagrams described by the left and right sub-trees.
A diagram is a \emph{cell} (as opposed to a composite) when its
top-level paste tree is a single $\mathsf{Leaf}$.

Paste trees serve two purposes. First, they determine how
$\mathsf{apply}$ traverses a diagram when applying a partial map
(\cref{sec:data:pmap}). Second, they distinguish structurally
identical diagrams that were built by different paste sequences,
which matters for the interpreter's notion of equality.

\paragraph{Operations.}
\begin{itemize}
  \item \textbf{Cell construction.}
    $\CellOf{t}{\mathit{cd}}$ constructs the diagram consisting of a
    single cell with tag $t$ and cell data $\mathit{cd}$.
    For a 0-cell, the shape is a single point. For a cell of
    dimension $k > 0$ with boundaries $\CBdy{d^-}{d^+}$,
    the shape is built from the shapes of $d^-$ and $d^+$.

  \item \textbf{Boundary.}
    $\bdy{\pm}{k}{\Diag{D}}$ extracts the $(k)$-boundary
    by extracting the boundary of the underlying shape and pulling
    back labels and paste trees through the embedding.

  \item \textbf{Pasting.}
    $\Diag{D}_1 \paste{k} \Diag{D}_2$ pastes two diagrams at
    dimension $k$. This requires that the output $k$-boundary of
    $\Diag{D}_1$ is isomorphic to the input $k$-boundary of
    $\Diag{D}_2$ (\emph{pastability}). The resulting shape is a
    pushout of the two underlying ogposets along the shared boundary.

  \item \textbf{Normalisation.}
    $\mathsf{norm}(\Diag{D})$ normalises the underlying shape and
    permutes labels accordingly.

  \item \textbf{Equality and isomorphism.}
    Two diagrams are \emph{equal} if they have identical shapes and
    matching labels at every position. They are \emph{isomorphic} if
    there is an isomorphism of the underlying shapes that preserves
    labels. The interpreter's \texttt{assert} command checks
    isomorphism.
\end{itemize}

% ---------------------------------------------------------------------------
\subsection{Partial Maps}
\label{sec:data:pmap}
% ---------------------------------------------------------------------------

Abstractly, a map between regular directed complexes is an
order-preserving function that covariantly preserves the
$\omega$-categorical structure (\cref{sec:molecules:regular}).  The
interpreter represents such a map as a \emph{partial map} $\Map{F}$: a
finite map $\Tag \rightharpoonup (\CData, \Diag{D})$ sending source
tags to their cell data and image diagram.  The domain is indexed by
dimension for ordered iteration.  A Boolean \emph{cellular} flag
records whether every image is a single cell (top-level paste tree is a
$\mathsf{Leaf}$).

\paragraph{Operations.}
\begin{itemize}
  \item \textbf{Extend.}
    $\mathsf{extend}(\Map{F}, t, \mathit{cd}, \Diag{D})$ adds a
    new entry mapping tag $t$ (with cell data $\mathit{cd}$) to
    image $\Diag{D}$, after validating \emph{boundary compatibility}:
    for a cell of dimension $k > 0$ with boundaries
    $\CBdy{d^-}{d^+}$, the map must already be defined on all
    boundary tags, and applying $\Map{F}$ to $d^-$ and $d^+$ must
    yield diagrams isomorphic to the corresponding boundaries of
    $\Diag{D}$.

  \item \textbf{Apply.}
    $\Map{F}(\Diag{D})$ applies $\Map{F}$ to a diagram by
    recursing on its paste tree:
    \begin{itemize}
      \item For $\mathsf{Leaf}(t)$: return $\Map{F}(t)$ (the image
        diagram).
      \item For $\mathsf{Node}(k, L, R)$: recursively apply to $L$
        and $R$, then paste the results at dimension $k$.
    \end{itemize}
    When $\Map{F}$ is cellular, a fast path avoids the paste-tree
    recursion: since every image is a single cell, it suffices to
    relabel the diagram's tags in place.

  \item \textbf{Compose.}
    $\Map{G} \circ \Map{F}$ composes two partial maps by applying
    $\Map{G}$ to each image in $\Map{F}$'s table.

  \item \textbf{Identity.}
    $\idmap{\Complex}$ maps every generator tag in $\Complex$ to
    its own classifier diagram. It is always cellular.
\end{itemize}

% ===========================================================================
\section{Abstract Syntax}
\label{sec:syntax}
% ===========================================================================

We present the abstract syntax in layers. We simplify the concrete syntax
wherever it would introduce unnecessary detail: for instance, we omit span
annotations and flatten optional fields. The concrete syntax of
\textsf{alifib} can be recovered by consulting the parser.

% ---------------------------------------------------------------------------
\subsection{Programs, Blocks, and Instructions}
\label{sec:syntax:prog}
% ---------------------------------------------------------------------------

A program is a sequence of blocks. There are two kinds: \emph{type blocks}
(introduced by $\kwtype$) for declaring new types, diagrams, maps, and
module inclusions at the top level; and \emph{local blocks} (introduced by
$\mathsf{@}\Cpx$) for working within an existing type's complex.

\begin{figure}[t]
\fbox{\text{programs and blocks}}
\[
  \begin{array}{rcll}
    \Prog  & \Coloneqq & \overline{\Block}         \\[3pt]
    \Block & \Coloneqq & \kwtype\;\overline{\TyInst}
             & \text{type block} \\
           &           & \mathsf{@}\Cpx\;\overline{\LocInst}
             & \text{local block}
  \end{array}
\]

\fbox{\text{type-level instructions}}
\[
  \begin{array}{rcll}
    \TyInst & \Coloneqq & n \;\decl\; \Cpx
              & \text{type declaration} \\
            &           & \kwlet\;n = \Diagram
              & \text{diagram naming} \\
            &           & \kwlet\;n \mathbin{::} \Addr = \PMapDef
              & \text{map definition} \\
            &           & \kwlet\;\kwtotal\;n \mathbin{::} \Addr = \PMapDef
              & \text{total map definition} \\
            &           & \kwinclude\;n\;[\kwas\;n']
              & \text{module inclusion}
  \end{array}
\]

\fbox{\text{complex-body instructions}}
\[
  \begin{array}{rcll}
    \CInst & \Coloneqq & n                   & \text{0-cell} \\
           &           & n : \Diagram \to \Diagram & \text{higher cell} \\
           &           & \kwlet\;n = \Diagram & \text{diagram naming} \\
           &           & \kwlet\;[\kwtotal]\;n \mathbin{::} \Addr = \PMapDef
             & \text{map definition} \\
           &           & \kwinclude\;\Addr\;[\kwas\;n]
             & \text{type inclusion} \\
           &           & \kwattach\;n \mathbin{::} \Addr\;[\kwalong\;\PMapDef]
             & \text{type attachment}
  \end{array}
\]

\fbox{\text{local instructions}}
\[
  \begin{array}{rcll}
    \LocInst & \Coloneqq & \kwlet\;n = \Diagram
               & \text{diagram naming} \\
             &           & \kwlet\;[\kwtotal]\;n \mathbin{::} \Addr = \PMapDef
               & \text{map definition} \\
             &           & \kwassert\;\Diagram = \Diagram
               & \text{equality assertion}
  \end{array}
\]
\caption{Programs and instructions.}
\label{fig:syntax:prog}
\end{figure}

\Cref{fig:syntax:prog} gives the full grammar of programs and
instructions. All forms that define partial maps use the $\kwlet$ keyword
in the concrete syntax; the distinction between diagram bindings and map
definitions is syntactic (map definitions include a $\mathbin{::} \Addr$
annotation). The keyword $\kwtotal$ marks a map definition as total,
meaning the interpreter checks that every generator in the source complex
has an image.

The type declaration form $n \;\decl\; \Cpx$ introduces a new 0-dimensional
type with a complex body. The symbol $\decl$ is written \texttt{<<=} in
the concrete syntax. The interpreter rejects type declarations with
non-trivial boundaries: all types are 0-cells, and higher cells are
declared inside complex bodies.

% ---------------------------------------------------------------------------
\subsection{Complexes and Addresses}
\label{sec:syntax:cpx}
% ---------------------------------------------------------------------------

A complex reference identifies which type's complex to operate in:
\[
  \begin{array}{rcll}
    \Cpx  & \Coloneqq & \Addr
            & \text{resolved address} \\
          &           & \Addr\;\{ \overline{\CInst} \}
            & \text{address with inline body} \\[3pt]
    \Addr & \Coloneqq & \varepsilon \mid n_1.n_2.\cdots.n_k
  \end{array}
\]
The empty address $\varepsilon$ refers to the current module's anonymous root
type. A non-empty address is resolved by navigating module-domain maps
for all segments except the last, and then looking up the final segment
as a cell diagram (\cref{sec:complex:addr}).

% ---------------------------------------------------------------------------
\subsection{Diagrams}
\label{sec:syntax:diag}
% ---------------------------------------------------------------------------

\begin{figure}[t]
\fbox{\text{diagram expressions}}
\[
  \begin{array}{rcll}
    \Diagram & \Coloneqq & \DExpr_1 \;\cdots\; \DExpr_k
               & \text{implicit pasting} \\
             &           & \Diagram \paste{k} \DExpr_1 \;\cdots\; \DExpr_j
               & \text{explicit pasting at dimension $k$} \\[3pt]
    \DExpr   & \Coloneqq & \DComp
               & \text{component} \\
             &           & \DExpr.\DComp
               & \text{dot access} \\[3pt]
    \DComp   & \Coloneqq & n
               & \text{name lookup} \\
             &           & (\Diagram)
               & \text{parenthesised diagram} \\
             &           & (\PMapG)
               & \text{parenthesised map} \\
             &           & (\,\kwmap\;\PMapDef \mathbin{::} \Cpx\,)
               & \text{anonymous map} \\
             &           & \kwin \mid \kwout
               & \text{boundary accessors} \\
             &           & \kwhole
               & \text{hole}
  \end{array}
\]

\fbox{\text{partial map expressions}}
\[
  \begin{array}{rcll}
    \PMapDef  & \Coloneqq & \PMapG
                & \text{plain map} \\
              &           & \PMapExt
                & \text{extension map} \\[3pt]
    \PMapG    & \Coloneqq & \PMapBasic
                & \text{basic map} \\
              &           & \PMapBasic.\PMapG
                & \text{composition via dot} \\[3pt]
    \PMapBasic & \Coloneqq & n
                & \text{named map} \\
              &           & (\,\kwmap\;\PMapDef \mathbin{::} \Cpx\,)
                & \text{anonymous map} \\
              &           & (\PMapG)
                & \text{parenthesised} \\[3pt]
    \PMapExt  & \Coloneqq & [\,\overline{\Clause}\,]
                & \text{extension from empty} \\
              &           & \PMapG\;[\,\overline{\Clause}\,]
                & \text{extension of prefix} \\[3pt]
    \Clause   & \Coloneqq & \Diagram \Rightarrow \Diagram
  \end{array}
\]
\caption{Diagram and partial map expressions.}
\label{fig:syntax:expr}
\end{figure}

\Cref{fig:syntax:expr} gives the grammar of expressions. There are two
kinds of pasting. \emph{Implicit pasting} (juxtaposition) pastes a
sequence of diagram expressions left to right, automatically computing the
paste dimension from the dimensions of the operands. \emph{Explicit
pasting} $\Diagram \paste{k} \overline{\DExpr}$ pastes at a specified
dimension $k$.

\paragraph{Name ambiguity.}
When a name $n$ is looked up in a complex, the interpreter checks the
diagram namespace first, then the map namespace. A name found as a diagram
produces a diagram term; a name found as a map produces a map term
carrying both the map and a reference to its source complex. This
overloading is intentional: the dot operator~(.) behaves differently
depending on whether its left operand is a diagram or a map
(\cref{sec:diag:dot}).

\paragraph{Holes.}
The hole symbol $\kwhole$ (written \texttt{?} in concrete syntax) is not
a first-class value. When the interpreter encounters a hole, it records
diagnostic information (including boundary context inferred from adjacent
expressions) and aborts diagram evaluation. Holes exist to support
interactive development.

% ===========================================================================
\section{Interpreter State}
\label{sec:state}
% ===========================================================================

\Cref{fig:domains} lists the semantic domains used in the formal
rules.  All are finite; the interpreter builds them incrementally.

\begin{figure}[t]
\fbox{\text{semantic domains}}

\medskip
\begin{tabular}{@{}l@{\;\;}c@{\;\;}l@{\qquad}l@{}}
$\State$    & $\in$ & $\mathbf{State}$
  & global interpreter states \\
$\Complex$  & $\in$ & $\mathbf{Cpx}$
  & complexes (namespaces) \\
$\Diag{D}$  & $\in$ & $\mathbf{Diag}$
  & diagrams (labelled ogposets with paste trees) \\
$\Map{F}$   & $\in$ & $\mathbf{PMap}$
  & partial maps (tag-indexed tables) \\
$\GId$      & $\in$ & $\GIds = \mathbb{N}$
  & global identifiers \\
$n$         & $\in$ & $\Namespace$
  & names (identifier strings) \\
$t$         & $\in$ & $\Tag = \TagG{\GId} \mid \TagL{n}$
  & tags (global or local) \\
$\mathit{cd}$ & $\in$ & $\CData = \CZero \mid \CBdy{\Diag{D}}{\Diag{D}}$
  & cell data \\
$m$         & $\in$ & $\ModId$
  & module identifiers (canonical file paths) \\
            & $\in$ & $\mathbf{MapDom} = \TyDom(\GIds) \mid \ModDom(\ModId)$
  & map domain annotations \\
\end{tabular}
\caption{Semantic domains and metavariable conventions.}
\label{fig:domains}
\end{figure}

The interpreter maintains a global state $\State \in \mathbf{State}$
consisting of three finite maps:
\[
  \State = \langle \cells, \types, \modules \rangle
\]

\begin{itemize}
  \item $\cells : \GIds \rightharpoonup \mathbf{CData}$ maps
    global identifiers of \emph{cell generators} (those declared inside
    complex bodies) to their cell data.  (Dimensions are stored in a
    separate side-table for efficient dimension-based iteration.)
  \item $\types : \GIds \rightharpoonup (\mathbf{CData} \times \mathbf{Cpx})$ maps
    global identifiers of \emph{type declarations} to their cell data and
    associated complex (namespace).  A type declaration
    (\texttt{T <<= \{...\}}) introduces a named regular directed complex;
    its cell data is always $\CZero$ (types are 0-dimensional), so the
    significant component is the complex.  The distinction from $\cells$
    is that types carry namespaces, while plain cell generators do not.
  \item $\modules : \ModId \rightharpoonup \mathbf{Cpx}$ maps canonical
    file paths to module-level complexes.
\end{itemize}

\noindent
Complexes are not independent of the state: every complex in the system
is stored inside $\State.\types$ or $\State.\modules$.  In the judgment
forms (\cref{fig:judgments}), the parameter~$\Complex$ is a
\emph{location}: it identifies which complex in the state the
instruction operates on.  For type-level and local instructions, the
location is determined by looking up the state (e.g.\
$\findmod{m}$ or $\findtype{\GId}.\Complex$), and any modifications to
the location are reflected in the returned~$\State'$.  For complex-body
instructions, the location is a \emph{working copy} threaded
separately---the caller is responsible for writing $\Complex'$ back
into the state when the body is complete.

% ---------------------------------------------------------------------------
\subsection{Global Identifiers and Tags}
\label{sec:state:gid}
% ---------------------------------------------------------------------------

The state uses \emph{global identifiers} $\GId \in \GIds$ to name
cells and types across the entire program.
We write $\GIds$ for the set of global identifiers (concretely
$\GIds = \mathbb{N}$); $\GId$ is a metavariable ranging over $\GIds$.
Identifiers are allocated by a
monotonic counter: $\fresh$ returns the next unused natural number. Every
globally-registered cell or type receives a unique $\GId$ at the point of
declaration.

The purpose of global identifiers is to provide a stable, unique identity
for each cell that is invariant under renaming. Two cells in different
complexes may bear the same human-readable name (e.g.\ \texttt{pt}) but
refer to entirely different generators; conversely, the \emph{same} cell
may appear in multiple complexes under different names (e.g.\ after an
$\kwinclude$). Global identifiers resolve this: they are the canonical
identity of a cell. In the Rust implementation, $\GIds$ is a wrapper around
\texttt{usize} allocated via a global atomic counter.

Every cell in a diagram is labelled not with a name but with a
\emph{tag} (\cref{fig:domains}).
Global tags $\TagG{\GId}$ reference cells registered in $\State.\cells$
or $\State.\types$. Local tags $\TagL{n}$ reference cells in a complex's
local cell store; they are used only inside complex bodies being
constructed and must not appear in persisted definitions. The reason for
having two kinds of tags is that local cells are temporary scaffolding
during complex construction: they exist in the working copy of a complex
but are not registered in the global state.

% ---------------------------------------------------------------------------
\subsection{Cell Data}
\label{sec:state:celldata}
% ---------------------------------------------------------------------------

Every cell has associated \emph{cell data} $\mathit{cd} \in \CData$
(\cref{fig:domains}) describing its boundary.
The symbol $\CZero$ denotes a 0-cell (no boundaries). The form
$\CBdy{d^{-}}{d^{+}}$ describes a cell of dimension $k > 0$ with
input boundary $d^{-}$ and output boundary $d^{+}$, both diagrams of
dimension $k - 1$.

% ---------------------------------------------------------------------------
\subsection{Complexes}
\label{sec:state:complex}
% ---------------------------------------------------------------------------

A \emph{complex} $\Complex$ is the interpreter's representation of a
(finite) regular directed complex, serving as the namespace associated
with a type or module.  It bundles four components:
\[
  \Complex = \langle \mathit{gens}, \mathit{diags}, \mathit{maps},
    \mathit{locals} \rangle
\]
\begin{itemize}
  \item $\mathit{gens} : \Namespace \rightharpoonup
    (\mathbf{Tag} \times \mathbb{N} \times \mathbf{Diag})$
    maps generator names to their tag, dimension, and
    \emph{classifier diagram} (the diagram consisting of that single cell).
  \item $\mathit{diags} : \Namespace \rightharpoonup \mathbf{Diag}$ maps diagram names
    to diagram values (representations of pasting diagrams;
    see \cref{sec:data}).
  \item $\mathit{maps} : \Namespace \rightharpoonup
    (\mathbf{MapDom} \times \mathbf{PMap})$ maps partial map names to a domain
    annotation and a partial map value (representing a map of
    directed complexes; see \cref{sec:data}).
    The domain is either $\TyDom(\GIds)$ or $\ModDom(\ModId)$.
  \item $\mathit{locals} : \Namespace \rightharpoonup \mathbf{CData}$
    stores locally-scoped cells (populated only during complex body
    interpretation). As with $\cells$, dimensions are maintained in
    a separate index.
\end{itemize}
A complex also maintains a set of all names in use, to detect collisions
eagerly. When a generator is \emph{declared} in a complex body, it is
entered into both $\mathit{gens}$ and $\mathit{diags}$: the former
records structural metadata while the latter enables name resolution
during diagram evaluation. Generators imported via $\kwinclude$ or
$\kwattach$ are entered only into $\mathit{gens}$; they are accessed
through dot notation on the inclusion map rather than by direct name
lookup in $\mathit{diags}$.

% ---------------------------------------------------------------------------
\subsection{Terms}
\label{sec:state:terms}
% ---------------------------------------------------------------------------

Diagram expressions evaluate to \emph{terms}:
\[
  \TermD \Coloneqq \DTerm{\Diag{D}} \mid
  \MTerm{\Map{F}, \Complex_{\mathit{src}}}
\]
A diagram term $\DTerm{\Diag{D}}$ carries a diagram (representing a
pasting diagram in a directed complex).  A map
term $\MTerm{\Map{F}, \Complex_{\mathit{src}}}$ carries a partial map
$\Map{F}$ (representing a map of directed complexes) together with a
reference to its \emph{source complex}. The
source complex is essential: it determines the namespace in which further
dot-access into the map is resolved.

The reason the interpreter tracks terms rather than just diagrams is that
partial maps and diagrams share the same expression language. When the user
writes \texttt{F.x}, the interpreter must know that \texttt{F} is a map
in order to resolve \texttt{x} in the source complex of \texttt{F} (not
in the ambient complex). For example, if \texttt{F} is a map from a type
\texttt{Group} to the current type, then \texttt{F.mul} looks up
\texttt{mul} in \texttt{Group}'s complex and returns its image under
\texttt{F}. The term type makes this dispatch explicit.

% ---------------------------------------------------------------------------
\subsection{State Threading}
\label{sec:state:threading}
% ---------------------------------------------------------------------------

The state $\State$ is threaded through interpretation.
Instructions within a block are processed sequentially: each receives
the state produced by the previous.  Local blocks stop on first error;
type blocks and complex bodies process all instructions regardless of
errors, accumulating diagnostics.

\Cref{fig:judgments} lists all judgment forms used in the remainder of
the document.

\begin{figure}[t]
\fbox{\text{judgment forms}}

\medskip
\begin{tabular}{@{}l@{\quad}l@{}}
$\IntJ{\State}{\Complex}{\mathit{instr}}{\State'}$
  & Interpret instruction $\mathit{instr}$ in state $\State$ at
    location~$\Complex$; \\
  & yields updated state $\State'$. \\[6pt]
$\IntJ{\State}{\Complex}{\mathit{instr}}{(\State', \Complex')}$
  & Variant for complex-body instructions: also returns \\
  & the updated complex $\Complex'$. \\[6pt]
$\EvalD{\State}{\Complex}{\Diagram}{\Diag{D}}$
  & Evaluate diagram expression $\Diagram$ in state $\State$ at \\
  & location~$\Complex$; yields a diagram $\Diag{D}$. \\[6pt]
$\EvalT{\State}{\Complex}{e}{\TermD}$
  & Evaluate expression $e$ to a term: either \\
  & $\DTerm{\Diag{D}}$ (diagram) or $\MTerm{\Map{F}, \Complex_S}$ (map). \\[6pt]
$\EvalM{\State}{\Complex}{(\PMapDef, \Complex_S)}{\Map{F}}$
  & Evaluate partial map definition $\PMapDef$ in state $\State$ \\
  & at location~$\Complex$ with source complex $\Complex_S$; \\
  & yields a partial map $\Map{F}$. \\[6pt]
$\ResJ{\State}{\Cpx}{(\GId, \Complex)}$
  & Resolve complex reference $\Cpx$ in state $\State$; \\
  & yields a type identifier $\GId$ and its complex $\Complex$. \\[6pt]
$\AddrJ{\State}{\Addr}{\GId}$
  & Resolve address $\Addr$ in state $\State$; \\
  & yields the type identifier $\GId$ it refers to. \\[6pt]
$\AssertJ{\State}{\Complex}{\mathit{assertion}}$
  & Check assertion in state $\State$ at location~$\Complex$; \\
  & succeeds (no output) or raises an error.
\end{tabular}
\caption{Summary of judgment forms.  The state $\State$ and location
  $\Complex$ are inputs; the right-hand side of $\Downarrow$ (or
  $\checkmark$) is the output.}
\label{fig:judgments}
\end{figure}

In the Rust implementation, $\State$ is wrapped in \texttt{Arc<State>} and
mutated via \texttt{Arc::make\_mut}. This copy-on-write discipline is not
observable in the formal rules but explains why the implementation freely
clones contexts before potentially-failing operations.
% ===========================================================================
\section{Program Interpretation}
\label{sec:prog}
% ===========================================================================

\subsection{Module Initialisation}
\label{sec:prog:init}

When a module $m$ is first interpreted, its complex is initialised with an
anonymous root type:

\begin{mathpar}
  \inferrule*[lab=\textsc{Prog-Skip}]
    {\findmod{m}\dn}
    {\IntJ{\State}{\cdot}{\Prog(m)}{\State}}
  \and
  \inferrule*[lab=\textsc{Prog-Init}]
    {\findmod{m}\up \\
     \GId_r = \fresh \\
     D_r = \CellOf{\TagG{\GId_r}}{\CZero} \\\\
     \Complex_0 = [\varepsilon \mapsto D_r] \\
     \State_1 = \State[\types(\GId_r) \mapsto (\CZero, \emptyset),\;
                        \modules(m) \mapsto \Complex_0] \\\\
     \IntJ{\State_1}{\Complex_0}{\overline{\Block}}{\State'}}
    {\IntJ{\State}{\cdot}{\Prog(m)}{\State'}}
\end{mathpar}

\textsc{Prog-Skip} ensures idempotent module loading: if $m$ is already
loaded, its interpretation is skipped entirely. This prevents
double-interpretation when two modules include the same dependency, but
it also means that the order of includes can affect what state a module
sees.

\textsc{Prog-Init} allocates a fresh $\GId_r$ for an anonymous root type,
creates its cell diagram $D_r$, and registers both the type (with an
empty complex) and the module (with a complex containing $D_r$ under the
empty name $\varepsilon$). The root type and the module complex are
\emph{distinct} objects: the module complex holds the root generator
\emph{reference}, while the root type's complex will hold generators
declared by the first $\kwtype$ block that has no explicit address.

\subsection{Block Dispatch}
\label{sec:prog:block}

\begin{mathpar}
  \inferrule*[lab=\textsc{Blk-Type}]
    {\IntJ{\State}{\findmod{m}}{\overline{\TyInst}}{\State'}}
    {\IntJ{\State}{\cdot}{\kwtype\;\overline{\TyInst}}{\State'}}
  \and
  \inferrule*[lab=\textsc{Blk-Local}]
    {\ResJ{\State}{\Cpx}{(\GId_r, \Complex)} \\
     \IntJ{\State}{\Complex}{\overline{\LocInst}}{\State'}}
    {\IntJ{\State}{\cdot}{\mathsf{@}\Cpx\;\overline{\LocInst}}{\State'}}
\end{mathpar}

A type block interprets its instructions within the module complex. A local
block first resolves its complex reference to a namespace
$(\GId_r, \Complex)$ and then interprets local instructions within that
complex.

% ===========================================================================
\section{Type-Level Instructions}
\label{sec:type}
% ===========================================================================

\subsection{Type Declaration}
\label{sec:type:gen}

\begin{mathpar}
  \inferrule*[lab=\textsc{TyDecl}]
    {n \notin \mathit{used}(\findmod{m}) \\
     \ResJ{\State}{\Cpx}{(\_, \Complex_{\mathit{def}})} \\
     \GId_T = \fresh \\
     D_T = \CellOf{\TagG{\GId_T}}{\CZero} \\\\
     F_{\mathit{id}} = \idmap{\Complex_{\mathit{def}}} \\
     \Complex'_{\mathit{def}} = \Complex_{\mathit{def}}
       [\mathit{maps}(n) \mapsto (\TyDom(\GId_T), F_{\mathit{id}})] \\\\
     \State' = \State[\types(\GId_T) \mapsto
       (\CZero, \Complex'_{\mathit{def}}),\;
       \modules(m)(n) \mapsto D_T]}
    {\IntJ{\State}{\findmod{m}}{n \;\decl\; \Cpx}{\State'}}
\end{mathpar}

A type declaration resolves the complex body $\Cpx$ (interpreting any cell
generators, includes, and attaches inside), allocates a fresh $\GId_T$,
constructs the identity map $F_{\mathit{id}}$ on the definition complex,
and adds this identity map to the type's own complex under its own name.
The classifier diagram $D_T$ is added to both the generator and diagram
namespaces of the module complex.

\paragraph{The self-referencing identity map.}
The identity map is added to the type's own complex under the type's name.
This is how a type ``refers to itself'': inside the complex of a type
$T$, the name $T$ resolves to the identity inclusion map. This is
essential for constructions like $\kwattach\;S \mathbin{::} \textrm{Semigroup}\;\kwalong\;[\textrm{Set} \Rightarrow \textrm{Set}]$, where the right-hand \textrm{Set}
refers to the current type's own identity.

% ---------------------------------------------------------------------------
\subsection{Diagram Naming}
\label{sec:type:let}
% ---------------------------------------------------------------------------

\begin{mathpar}
  \inferrule*[lab=\textsc{TyLet}]
    {\EvalD{\State}{\findmod{m}}{\Diagram}{D}}
    {\IntJ{\State}{\findmod{m}}{\kwlet\;n = \Diagram}%
          {\State[\modules(m).\mathit{diags}(n) \mapsto D]}}
\end{mathpar}

The diagram expression is evaluated and bound in the module complex.
Unlike complex-body and local-block let-bindings, the Rust
implementation does \emph{not} check name uniqueness at the type level:
a duplicate name silently overwrites the previous binding.

% ---------------------------------------------------------------------------
\subsection{Map Definition}
\label{sec:type:def}
% ---------------------------------------------------------------------------

\begin{mathpar}
  \inferrule*[lab=\textsc{TyMap}]
    {\AddrJ{\State}{\Addr}{\GId_S} \\
     \Complex_S = \findtype{\GId_S}.\Complex \\
     \EvalM{\State}{\findmod{m}}{(\PMapDef, \Complex_S)}{F}}
    {\IntJ{\State}{\findmod{m}}{\kwlet\;n \mathbin{::} \Addr = \PMapDef}%
          {\State[\modules(m).\mathit{maps}(n) \mapsto
            (\TyDom(\GId_S), F)]}}
\end{mathpar}

The address resolves to the source type $\GId_S$. The map definition is
evaluated with $\Complex_S$ as the source complex. For $\kwtotal$ maps, an
additional post-hoc check verifies that every generator in $\Complex_S$
has an image; the map is returned even if the check fails (errors
accumulate).

As with \textsc{TyLet}, name uniqueness is not checked at the type level.

% ---------------------------------------------------------------------------
\subsection{Module Inclusion}
\label{sec:type:import}
% ---------------------------------------------------------------------------

\begin{mathpar}
  \inferrule*[lab=\textsc{TyInclude}]
    {\mathit{alias} = n'\text{ if given, else }n \\
     \mathit{alias} \notin \mathit{used}(\findmod{m}) \\
     m' = \mathsf{resolve}(m, n) \\
     \IntJ{\State}{\cdot}{\Prog(m')}{\State_1} \\\\
     \Complex_{m'} = \State_1.\modules(m') \\
     F_{\mathit{id}} = \idmap{\Complex_{m'}} \\
     \mathit{gs} = \{ (\mathit{alias}.n_g, t_g, k_g, D_g) \mid
       (n_g, (t_g, k_g, D_g)) \in \Complex_{m'}.\mathit{gens},\;
       n_g \neq \varepsilon \} \\\\
     \State' = \State_1[
       \modules(m).\mathit{gens} \mathbin{{+}{=}} \mathit{gs},\;
       \modules(m).\mathit{maps}(\mathit{alias}) \mapsto
         (\ModDom(m'), F_{\mathit{id}})]}
    {\IntJ{\State}{\findmod{m}}{\kwinclude\;n\;[\kwas\;n']}{\State'}}
\end{mathpar}

Module inclusion resolves $n$ to a canonical path $m'$, recursively
interprets $m'$ (which may transitively include further modules), and then
merges the included module's generators into the current module's complex.
The alias defaults to $n$ when no $\kwas$ clause is given; in either case,
the alias is checked for uniqueness in the current module.
Generators are prefixed with the alias and
duplicates---identified by tag, not by name---are silently skipped. This
is how diamond dependencies are handled: if two imports bring in the same
cell (same $\GId$), only the first copy is kept.

The state $\State_1$ from the recursive interpretation is carried forward
as the base for further modifications, ensuring that all types, cells, and
modules registered during the recursive call are visible.

% ===========================================================================
\section{Complex Resolution}
\label{sec:complex}
% ===========================================================================

Complex references resolve to namespaces $(\GId, \Complex)$:

\begin{mathpar}
  \inferrule*[lab=\textsc{Cpx-Root}]
    {\findmod{m}.\mathit{gens}(\varepsilon) = (\TagG{\GId_r}, \_, \_) \\
     \Complex = \findtype{\GId_r}.\Complex}
    {\ResJ{\State}{\varepsilon}{(\GId_r, \Complex)}}
  \and
  \inferrule*[lab=\textsc{Cpx-Addr}]
    {\AddrJ{\State}{\Addr}{\GId} \\
     \Complex = \findtype{\GId}.\Complex}
    {\ResJ{\State}{\Addr}{(\GId, \Complex)}}
  \and
  \inferrule*[lab=\textsc{Cpx-Block}]
    {\ResJ{\State}{\Addr}{(\GId, \Complex_0)} \\
     \IntJ{\State}{\Complex_0}{\overline{\CInst}}{(\State', \Complex')}}
    {\ResJ{\State}{\Addr\;\{\overline{\CInst}\}}{(\GId, \Complex')}}
\end{mathpar}

\textsc{Cpx-Root} handles the empty address by looking up the anonymous root
generator in the module complex. \textsc{Cpx-Addr} resolves a non-empty
address to a type's complex. \textsc{Cpx-Block} first resolves the
address, then interprets the body instructions, threading both state and
location; the body may add generators, diagrams, maps, includes, and
attaches. State changes from the body persist to the outer context.

\subsection{Address Resolution}
\label{sec:complex:addr}

\begin{mathpar}
  \inferrule*[lab=\textsc{Addr-Leaf}]
    {\Complex.\mathit{diags}(n) = D \\
     D\;\text{is a cell} \\
     \mathit{top}(D) = \TagG{\GId}}
    {\AddrJ{\State}{n}{\GId}}
  \and
  \inferrule*[lab=\textsc{Addr-Step}]
    {\Complex.\mathit{maps}(n_1) = (\ModDom(m'), \_) \\
     \Complex' = \findmod{m'} \\
     \AddrJ{\State[\Complex']}
       {n_2.\cdots.n_k}{\GId}}
    {\AddrJ{\State}{n_1.n_2.\cdots.n_k}{\GId}}
\end{mathpar}

Address resolution navigates the \emph{map} namespace, not the generator
namespace. All segments except the last must be names of maps with module
domains; each step follows the module reference to its complex. The final
segment is looked up in the diagram namespace and must be a single cell.

\paragraph{Why maps, not generators?}
The map namespace is used because module inclusions register an identity
map under the alias name. This map's domain records which module (or type)
the inclusion came from, and its source complex provides the namespace for
further resolution.  Using maps rather than generators allows the same
resolution mechanism to work for both module paths
(\texttt{Theory.Equation}) and type paths (\texttt{F.S}).

% ===========================================================================
\section{Complex Body Instructions}
\label{sec:cinstr}
% ===========================================================================

Complex body instructions are interpreted in a mode $\mu \in \{\ModeG, \ModeL\}$.
In $\ModeG$ mode, new generators receive global tags and are registered in
$\State.\cells$. In $\ModeL$ mode, they receive local tags and are stored
in the complex's local cell store.

\subsection{Cell Generator}
\label{sec:cinstr:gen}

\begin{mathpar}
  \inferrule*[lab=\textsc{CGen-Zero}]
    {n \notin \mathit{used}(\Complex) \\
     \mu = \ModeG \implies
       \GId = \fresh,\;
       t = \TagG{\GId},\;
       \State' = \State[\cells(\GId) \mapsto \CZero] \\
     \mu = \ModeL \implies
       t = \TagL{n},\;
       \State' = \State \\\\
     D = \CellOf{t}{\CZero} \\
     \Complex' = \Complex[\mathit{gens}(n) \mapsto (t, 0, D),\;
       \mathit{diags}(n) \mapsto D]}
    {\IntJ{\State}{\Complex}{n}{\State', \Complex'}}
  \and
  \inferrule*[lab=\textsc{CGen-Cell}]
    {n \notin \mathit{used}(\Complex) \\
     \EvalD{\State}{\Complex}{\Diagram^-}{D^-} \\
     \EvalD{\State}{\Complex}{\Diagram^+}{D^+} \\
     D^-,\, D^+\;\text{round} \\\\
     \mathit{cd} = \CBdy{D^-}{D^+} \qquad
     k = \dim(D^-) + 1 \\
     \mu = \ModeG \implies
       \GId = \fresh,\;
       t = \TagG{\GId},\;
       \State' = \State[\cells(\GId) \mapsto \mathit{cd}] \\
     \mu = \ModeL \implies
       t = \TagL{n},\;
       \State' = \State \\\\
     D = \CellOf{t}{\mathit{cd}} \\
     \Complex' = \Complex[\mathit{gens}(n) \mapsto (t, k, D),\;
       \mathit{diags}(n) \mapsto D]}
    {\IntJ{\State}{\Complex}{n : \Diagram^- \to \Diagram^+}{\State', \Complex'}}
\end{mathpar}

\textsc{CGen-Zero} declares a 0-cell (a bare name with no boundary).
\textsc{CGen-Cell} declares a higher cell: the source and target diagrams
$\Diagram^-$ and $\Diagram^+$ are evaluated in the current complex, and
their roundness is checked (this ensures the resulting closure is an
atom; see \cref{sec:molecules:regular}).

The cell diagram $D = \CellOf{t}{\mathit{cd}}$ is constructed by taking
the pushout of the input and output boundary shapes into a common lower
boundary, then adding one new top-dimensional cell (\cref{sec:data}).
In $\ModeL$ mode, the cell data is also added to
$\Complex'.\mathit{locals}(n)$.

\subsection{Complex-Body Let and Def}
\label{sec:cinstr:let}

\begin{mathpar}
  \inferrule*[lab=\textsc{CLet}]
    {n \notin \mathit{used}(\Complex) \\
     \EvalD{\State}{\Complex}{\Diagram}{D}}
    {\IntJ{\State}{\Complex}{\kwlet\;n = \Diagram}
          {\State, \Complex[\mathit{diags}(n) \mapsto D]}}
  \and
  \inferrule*[lab=\textsc{CDef}]
    {n \notin \mathit{used}(\Complex) \\
     \AddrJ{\State}{\Addr}{\GId_S} \\
     \EvalM{\State}{\Complex}{(\PMapDef, \Complex_S)}{F}}
    {\IntJ{\State}{\Complex}{\kwlet\;n \mathbin{::} \Addr = \PMapDef}
          {\State, \Complex[\mathit{maps}(n) \mapsto (\TyDom(\GId_S), F)]}}
\end{mathpar}

All \texttt{let} forms---type-level, complex-body, and local---check name
uniqueness.

\subsection{Include}
\label{sec:cinstr:include}

\begin{mathpar}
  \inferrule*[lab=\textsc{CInclude}]
    {\AddrJ{\State}{\Addr}{\GId} \\
     \Complex_S = \findtype{\GId}.\Complex \\
     n \notin \mathit{used}(\Complex) \\
     F_{\mathit{id}} = \idmap{\Complex_S} \\\\
     \mathit{gs} = \{ (n.n_g, t_g, k_g, D_g) \mid
       (n_g, (t_g, k_g, D_g)) \in \Complex_S.\mathit{gens},\;
       t_g \notin \Complex \} \\
     \Complex' = \Complex[\mathit{gens} \mathbin{{+}{=}} \mathit{gs},\;
       \mathit{maps}(n) \mapsto (\TyDom(\GId), F_{\mathit{id}})]}
    {\IntJ{\State}{\Complex}{\kwinclude\;\Addr\;[\kwas\;n]}%
          {\State, \Complex'}}
\end{mathpar}

Include copies generators from the source type's complex into the current
complex, prefixing names, and adds an identity map. Crucially, no fresh
identifiers are allocated: the included generators retain their original
tags. This means that the ``same'' cell can appear in multiple complexes
under different names, but it has one canonical identity.

If no alias is given and the name would contain a dot (indicating a
non-local type), the interpreter raises an error: inclusion of non-local
types requires an explicit alias.

\subsection{Attach}
\label{sec:cinstr:attach}

\begin{mathpar}
  \inferrule*[lab=\textsc{CAttach}]
    {\AddrJ{\State}{\Addr}{\GId_A} \\
     \Complex_A = \findtype{\GId_A}.\Complex \\
     n \notin \mathit{used}(\Complex) \\
     F_0 = \text{(evaluate along map, or } \emptyset \text{ if omitted)} \\\\
     \text{for each } (n_g, (t_g, k_g, \_)) \in \Complex_A.\mathit{gens}
       \text{, by ascending } k_g\text{, with } F_i(t_g)\up\text{:} \\
     \quad \GId' = \fresh \qquad
     \mathit{cd}' = \St{\mathsf{bdy}}(F_i, t_g) \qquad
     D' = \CellOf{\TagG{\GId'}}{\mathit{cd}'} \\
     \quad F_{i+1} = F_i[t_g \mapsto D'] \qquad
     \State_{i+1} = \State_i[\cells(\GId') \mapsto \mathit{cd}'] \\\\
     \Complex' = \Complex[\mathit{gens} \mathbin{{+}{=}} \mathit{new},\;
       \mathit{maps}(n) \mapsto (\TyDom(\GId_A), F_{\mathit{final}})]}
    {\IntJ{\State}{\Complex}{\kwattach\;n \mathbin{::} \Addr\;
       [\kwalong\;\PMapDef]}{\State', \Complex'}}
\end{mathpar}

\noindent
Here $\St{\mathsf{bdy}}(F, t)$ is $\CZero$ if the cell is 0-dimensional, and
$\CBdy{\apply{F}{d^{-}}}{\apply{F}{d^{+}}}$ otherwise (where
$\CBdy{d^{-}}{d^{+}}$ is the cell data of $t$).

Attachment differs from inclusion in a fundamental way: it creates
\emph{fresh} cells for every generator not already in the ``along'' map.
The ``along'' map specifies which generators of the attached type should
be identified with existing generators in the current complex (e.g.,
sharing a common base type), while every other generator gets a fresh copy.
Generators are processed in ascending dimension order, ensuring that when
we encounter an $n$-cell, all of its boundary cells have already been
mapped (either by the ``along'' map or by a previous iteration step).

\paragraph{Attach vs.\ include.}
Include reuses existing tags and creates no new cells; the result is an
aliased view of the same generators. Attach creates genuinely new cells
with fresh $\GId$s and new tag identities. This is the mechanism for
instantiating a ``template'': attaching the same type twice produces two
independent copies.

% ===========================================================================
\section{Local Instructions}
\label{sec:local}
% ===========================================================================

Local instructions operate within a resolved namespace
$(\GId_r, \Complex)$. Their defining feature is that definitions are
\emph{persisted}: the interpreter writes them back into the type's complex
in the global state.

\subsection{Local Let and Def}
\label{sec:local:let}

\begin{mathpar}
  \inferrule*[lab=\textsc{LocLet}]
    {n \notin \mathit{used}(\Complex) \\
     \EvalD{\State}{\Complex}{\Diagram}{D} \\
     D\;\text{has no local tags}}
    {\IntJ{\State}{(\GId_r, \Complex)}{\kwlet\;n = \Diagram}%
          {\State[\types(\GId_r).\Complex.\mathit{diags}(n) \mapsto D]}}
  \and
  \inferrule*[lab=\textsc{LocDef}]
    {n \notin \mathit{used}(\Complex) \\
     \EvalM{\State}{\Complex}{(\PMapDef, \Complex_S)}{F} \\
     F\;\text{has no local tags}}
    {\IntJ{\State}{(\GId_r, \Complex)}
      {\kwlet\;n \mathbin{::} \Addr = \PMapDef}%
      {\State[\types(\GId_r).\Complex.\mathit{maps}(n) \mapsto
        (\TyDom(\GId_S), F)]}}
\end{mathpar}

Before persisting, the interpreter checks that the diagram or map
contains no locally-tagged cells. This ensures that definitions exported to
global state are self-contained. (This check is \emph{not} performed for
definitions inside complex bodies, which may legitimately refer to local
cells.)

The ``local'' in ``local block'' refers to the \emph{scope of operation}
(which type's complex we are working in), not to the lifetime of the
definition. Local-block definitions are global in duration.

\subsection{Assertions}
\label{sec:local:assert}

\begin{mathpar}
  \inferrule*[lab=\textsc{Assert-Diag}]
    {\EvalD{\State}{\Complex}{\Diagram_1}{D_1} \\
     \EvalD{\State}{\Complex}{\Diagram_2}{D_2} \\
     D_1 \cong D_2}
    {\AssertJ{\State}{\Complex}{\kwassert\;\Diagram_1 = \Diagram_2}}
  \and
  \inferrule*[lab=\textsc{Assert-Map}]
    {\EvalT{\State}{\Complex}{\Diagram_1}{\MTerm{F_1, \Complex_S}} \\
     \EvalT{\State}{\Complex}{\Diagram_2}{\MTerm{F_2, \Complex_S}} \\\\
     \forall (n, (t, \_, \_)) \in \Complex_S.\mathit{gens}{:}\;
       F_1(t)\dn \Leftrightarrow F_2(t)\dn \\
     \forall t\;\text{with } F_1(t)\dn{:}\;
       F_1(t) \cong F_2(t)}
    {\AssertJ{\State}{\Complex}{\kwassert\;\Diagram_1 = \Diagram_2}}
\end{mathpar}

Both sides must evaluate to terms of the same kind (both diagram or both
map). Diagram equality is checked via isomorphism of the underlying
ogposets with matching labels. Map equality is checked pointwise: for
each generator $t$ in the shared source complex, both maps must satisfy
$F_1(t)\dn \Leftrightarrow F_2(t)\dn$, and where both are defined, $F_1(t) \cong F_2(t)$.

\paragraph{Structural, not propositional.}
Assertions check \emph{definitional} (structural) equality---isomorphism
of the underlying ogposets with matching labels. There are no non-trivial
equalities: everything interesting is a \emph{rewrite} (a
higher-dimensional cell). For example, the unit law
$\kwassert\;\mathit{id}\;{}\alpha = \alpha$ will \emph{fail}, because
$\mathit{id}\;{}\alpha$ has an additional identity cell in its ogposet
that $\alpha$ alone does not. To express the unit law, one declares a
cell $\lambda : \mathit{id}\;{}\alpha \to \alpha$, which is itself a
rewrite witnessing the relationship.

However, middle-four interchange---the equation
$(\alpha \paste{0} \beta)(\gamma \paste{0} \delta) =
(\alpha\gamma) \paste{0} (\beta\delta)$---\emph{is} structural: both
sides produce isomorphic ogposets.

% ===========================================================================
\section{Diagram Evaluation}
\label{sec:diag}
% ===========================================================================

\subsection{Implicit Pasting}
\label{sec:diag:juxt}

\begin{mathpar}
  \inferrule*[lab=\textsc{Diag-One}]
    {\EvalT{\State}{\Complex}{\DExpr}{\DTerm{D}}}
    {\EvalD{\State}{\Complex}{\DExpr}{D}}
  \and
  \inferrule*[lab=\textsc{Diag-Juxt}]
    {\EvalD{\State}{\Complex}{\DExpr_1\;\cdots\;\DExpr_{k-1}}{D_L} \\
     \EvalD{\State}{\Complex}{\DExpr_k}{D_R} \\\\
     j = \min(\dim(D_L), \dim(D_R)) - 1 \\
     D = D_L \paste{j} D_R}
    {\EvalD{\State}{\Complex}{\DExpr_1\;\cdots\;\DExpr_k}{D}}
\end{mathpar}

Juxtaposition is left-associative. The Rust implementation evaluates
implicit pasting left to right: the first expression is evaluated, then
each subsequent expression is evaluated and pasted onto the accumulator.
This differs from explicit pasting, which evaluates the right side before
the left (\cref{sec:diag:paste}). The paste dimension is
$\min(\dim(D_L), \dim(D_R)) - 1$. For two 1-cells this gives $j = 0$
(horizontal composition); for two 2-cells, $j = 1$ (vertical
composition).

\subsection{Explicit Pasting}
\label{sec:diag:paste}

\begin{mathpar}
  \inferrule*[lab=\textsc{Diag-Paste}]
    {\EvalD{\State}{\Complex}{\overline{\DExpr}_R}{D_R} \\
     \EvalD{\State}{\Complex}{\Diagram_L}{D_L} \\
     \bdy{+}{k}{D_L} \cong \bdy{-}{k}{D_R} \\
     D = D_L \paste{k} D_R}
    {\EvalD{\State}{\Complex}{\Diagram_L \paste{k}
      \overline{\DExpr}_R}{D}}
\end{mathpar}

The right-hand side is evaluated before the left-hand side, matching
the OCaml reference implementation's evaluation order.
The paste is defined only when $D_L$'s output $k$-boundary
matches $D_R$'s input $k$-boundary. The result is computed as a pushout
of the two boundary embeddings, merging labels and paste trees.

\subsection{Component Evaluation}
\label{sec:diag:comp}

\begin{mathpar}
  \inferrule*[lab=\textsc{Comp-Name-D}]
    {\Complex.\mathit{diags}(n) = D}
    {\EvalT{\State}{\Complex}{n}{\DTerm{D}}}
  \and
  \inferrule*[lab=\textsc{Comp-Name-M}]
    {\Complex.\mathit{diags}(n)\up \\
     \Complex.\mathit{maps}(n) = (\mathit{dom}, F) \\
     \Complex_S = \St{\mathsf{src}}(\State, \mathit{dom})}
    {\EvalT{\State}{\Complex}{n}{\MTerm{F, \Complex_S}}}
\end{mathpar}

\begin{mathpar}
  \inferrule*[lab=\textsc{Comp-Paren}]
    {\EvalD{\State}{\Complex}{\Diagram}{D}}
    {\EvalT{\State}{\Complex}{(\Diagram)}{\DTerm{D}}}
  \and
  \inferrule*[lab=\textsc{Comp-PMap}]
    {\EvalM{\State}{\Complex}{(\PMapG, \Complex)}{F, \Complex_S}}
    {\EvalT{\State}{\Complex}{(\PMapG)}{\MTerm{F, \Complex_S}}}
\end{mathpar}

\begin{mathpar}
  \inferrule*[lab=\textsc{Comp-AnonMap}]
    {\ResJ{\State}{\Cpx}{(\_, \Complex_T)} \\
     \EvalM{\State}{\Complex_T}{(\PMapDef, \Complex)}{F}}
    {\EvalT{\State}{\Complex}{(\,\kwmap\;\PMapDef :: \Cpx\,)}
      {\MTerm{F, \Complex}}}
  \and
  \inferrule*[lab=\textsc{Comp-Bd}]
    {\strut}
    {\EvalT{\State}{\Complex}{\kwin}{\mathsf{Bd}(\text{in})}}
\end{mathpar}

Name lookup checks diagrams first, then maps.

\paragraph{Anonymous maps in diagram context.}
Anonymous maps resolve $\Cpx$ to obtain $\Complex_T$, which
becomes the evaluation context for clause right-hand sides (images).
Clause left-hand sides are resolved in the \emph{ambient} complex
$\Complex$, which also becomes the returned map's source. In other words,
the $\mathbin{::}\Cpx$ annotation in a diagram-level anonymous map
specifies where images live, not where sources come from. This differs
from the convention in named map definitions ($\kwlet\;n \mathbin{::}
\Addr = \ldots$), where $\Addr$ identifies the source type.

The source complex for
$\St{\mathsf{src}}(\State, \mathit{dom})$ is
$\findtype{\GId}.\Complex$ if $\mathit{dom} = \TyDom(\GId)$, or
$\findmod{m}$ if $\mathit{dom} = \ModDom(m)$.

\subsection{Dot Access}
\label{sec:diag:dot}

The dot operator has three semantically distinct behaviours:

\begin{mathpar}
  \inferrule*[lab=\textsc{Dot-Bd}]
    {\EvalT{\State}{\Complex}{\DExpr}{\DTerm{D}} \\
     k = \max(\dim(D) - 1, 0) \\
     D' = \bdy{s}{k}{D}}
    {\EvalT{\State}{\Complex}{\DExpr.\mathsf{Bd}(s)}{\DTerm{D'}}}
\end{mathpar}

\begin{mathpar}
  \inferrule*[lab=\textsc{Dot-Apply}]
    {\EvalT{\State}{\Complex}{\DExpr}{\MTerm{F, \Complex_S}} \\
     \EvalT{\State}{\Complex_S}{\DComp}{\DTerm{D}} \\
     D' = \apply{F}{D}}
    {\EvalT{\State}{\Complex}{\DExpr.\DComp}{\DTerm{D'}}}
\end{mathpar}

\begin{mathpar}
  \inferrule*[lab=\textsc{Dot-Compose}]
    {\EvalT{\State}{\Complex}{\DExpr}{\MTerm{F, \Complex_S}} \\
     \EvalT{\State}{\Complex_S}{\DComp}{\MTerm{G, \Complex_{S'}}} \\
     H = \compose{F}{G}}
    {\EvalT{\State}{\Complex}{\DExpr.\DComp}
      {\MTerm{H, \Complex_{S'}}}}
\end{mathpar}

\textsc{Dot-Bd}: extracting the codimension-1 boundary of a diagram.

\textsc{Dot-Apply}: the key rule. The right-hand component is evaluated
in the source complex $\Complex_S$ of the map, not in the ambient
complex $\Complex$. This \emph{context switch} is how \texttt{F.x}
resolves \texttt{x} in $F$'s domain rather than in the surrounding
namespace. The result is the image of $D$ under $F$.

\textsc{Dot-Compose}: when both sides are maps, compose them. The
source complex of the result is $\Complex_{S'}$ (the inner map's source).

\paragraph{Dot on diagram with a non-boundary component} produces an error.
Similarly, a boundary accessor on a map term produces an error. These
cases are intentionally not given rules: they signal a type error in the
user's program.

% ===========================================================================
\section{Partial Map Evaluation}
\label{sec:pmap}
% ===========================================================================

Map expressions evaluate relative to a source complex $\Complex_S$
(the domain of the map being defined). We write the judgment as
$\EvalM{\State}{\Complex}{(\PMapDef, \Complex_S)}{F}$.

\subsection{Basic Lookup and Composition}
\label{sec:pmap:basic}

\begin{mathpar}
  \inferrule*[lab=\textsc{PM-Name}]
    {\Complex.\mathit{maps}(n) = (\mathit{dom}, F) \\
     \Complex_S = \St{\mathsf{src}}(\State, \mathit{dom})}
    {\EvalM{\State}{\Complex}{n}{\MC{F}{\Complex_S}}}
  \and
  \inferrule*[lab=\textsc{PM-Dot}]
    {\EvalM{\State}{\Complex}{\PMapBasic}{\MC{F}{\Complex_S}} \\
     \EvalM{\State}{\Complex_S}{\PMapG}{\MC{G}{\Complex_{S'}}} \\
     H = \compose{F}{G}}
    {\EvalM{\State}{\Complex}{\PMapBasic.\PMapG}{\MC{H}{\Complex_{S'}}}}
\end{mathpar}

Map composition via dot follows the same pattern as diagram dot access:
the right side is resolved in the left side's source complex.

\subsection{Extension Maps}
\label{sec:pmap:ext}

\begin{mathpar}
  \inferrule*[lab=\textsc{PM-Ext}]
    {F_0 = \text{(prefix map, or } \emptyset \text{)} \\
     \text{for each clause } \Diagram_i \Rightarrow \Diagram'_i\text{:} \\
     \quad \EvalD{\State}{\Complex_S}{\Diagram_i}{D_i} \qquad
     \EvalD{\State}{\Complex}{\Diagram'_i}{D'_i} \\
     \quad F_{i+1} = \smartext(F_i, \Complex_S, \Complex, D_i, D'_i)}
    {\EvalM{\State}{\Complex}{(\PMapExt, \Complex_S)}{F_{\mathit{final}}}}
\end{mathpar}

In each clause, the left-hand side is evaluated in the source complex
$\Complex_S$ and the right-hand side in the current location $\Complex$.
This is because each clause maps a cell of the source type to a diagram
in the ambient complex. Clauses are processed sequentially: each extends
the map built by the previous, so later clauses see earlier mappings.

\subsection{The Smart Extension Algorithm}
\label{sec:pmap:smart}

The function $\smartext(F, \Complex_S, \Complex_T, D_s, D_t)$ extends
a partial map $F$ with a mapping from a source cell $D_s$ to a target
diagram $D_t$, automatically inferring mappings for unmapped boundary
cells:

\begin{enumerate}
  \item \textbf{Cell check.} $D_s$ must be a single cell. Extract its
    tag $t$.
  \item \textbf{Idempotence.} If $F(t)\dn$ and $F(t) \cong D_t$,
    return $F$ unchanged. If $F(t)\dn$ but $F(t) \not\cong D_t$,
    raise an error.
  \item \textbf{Boundary collection.} Look up the cell data for $t$
    (checking $\State.\cells$ first, then $\State.\types$ as fallback,
    since types also carry cell data). If
    it has boundaries $\CBdy{d^{-}}{d^{+}}$, collect all top-dimensional
    tags $t_b$ in both for which $F(t_b)\up$.
  \item \textbf{Recursive extension.} For each such $t_b$
    with sign $s$:
    \begin{enumerate}
      \item Compute $D_b = \bdy{s}{\dim-1}{D_t}$ (the corresponding
        boundary of the \emph{target}).
      \item If the source boundary is a single cell, recurse.
      \item If the source boundary is composite, find an isomorphism
        between the source and target boundary shapes, use it to
        identify which target tag corresponds to $t_b$, then recurse.
    \end{enumerate}
  \item \textbf{Validation.} Call $\mathsf{PMap{:}{:}extend}$ to add
    $t \mapsto D_t$, which checks boundary compatibility.
\end{enumerate}

\paragraph{Isomorphism-based inference.}
When the source boundary is composite, the algorithm must determine which
cell in the target boundary corresponds to each unmapped source cell.
It normalises both boundary ogposets, finds a bijection between their
cells, and uses this to transfer labels. If a source tag appears at
multiple positions in the boundary, the algorithm checks that all
corresponding target positions carry the same label. This is the mechanism
that makes terse definitions like \texttt{[merge => split]} automatically
infer boundary mappings: the user maps only the top-dimensional cells,
and the algorithm deduces where the boundary cells must go by matching
boundary shapes.

\subsection{Map-to-Map Assignment}
\label{sec:pmap:mapassign}

When a clause maps a map term to another map term (both $\MTerm{\cdot}$),
the interpreter processes it as a pointwise extension:
\begin{enumerate}
  \item Verify that both maps share the same source complex.
  \item Iterate over generators of the shared source, sorted by dimension.
  \item For each generator $t$ with $F_L(t)\dn$ and $F_R(t)\dn$: if the
    left image is a cell, call $\smartext$ to map the left image to the
    right image. If the left image is composite, check that all its
    labels $t'$ satisfy $F(t')\dn$ (but do not recurse).
  \item If $F_L(t)\dn \not\Leftrightarrow F_R(t)\dn$ for any $t$, raise an error.
\end{enumerate}

This handles constructions like \texttt{[Dom => Ob, Cod => Ob]} where
\texttt{Dom} and \texttt{Cod} resolve to maps.

% ===========================================================================
\section{Map Application and Composition}
\label{sec:apply}
% ===========================================================================

The operation $\apply{F}{D}$ applies a partial map to a diagram. There
are two paths:

\begin{itemize}
  \item \textbf{Cellular.} If $F$ is cellular (all images are single
    cells), simply remap each label tag in $D$ to the corresponding image
    tag. The ogposet shape is shared. This is the common case for identity
    maps and inclusion maps.
  \item \textbf{General.} Follow the paste tree of $D$ recursively. For a
    $\mathsf{Leaf}(t)$, return $F(t)$. For a $\mathsf{Node}(k, L, R)$,
    recursively apply to $L$ and $R$, then paste the results at dimension
    $k$.
\end{itemize}

The identity map $\idmap{\Complex}$ maps every generator's tag to its own
classifier diagram. It is always cellular and therefore enjoys the fast
application path.

Map composition $\compose{F}{G}$ is computed eagerly: for each entry
$t \mapsto D$ in $G$'s table, the result maps $t$ to $\apply{F}{D}$.

% ===========================================================================
\section{Summary of Subtleties}
\label{sec:pain}
% ===========================================================================

\begin{enumerate}
  \item \textbf{Name ambiguity.} A name may resolve to a diagram or a
    map. Diagrams are checked first. The dot operator dispatches on the
    left operand's kind.

  \item \textbf{Evaluation order.} Implicit pasting is left-associative
    and evaluates left to right.
    Explicit pasting ($\paste{k}$) evaluates the right side first,
    matching the OCaml reference implementation.

  \item \textbf{Source complex switch.} Dotting into a map evaluates the
    right component in the map's source complex, not the ambient complex.
    This is the mechanism behind \texttt{F.x}.

  \item \textbf{Local block persistence.} Definitions in local blocks are
    persisted to global state. ``Local'' refers to scope of operation, not
    lifetime.

  \item \textbf{No local tags in persisted definitions.} Local blocks
    check for local tags before persisting. Complex bodies do not.

  \item \textbf{Attach creates fresh cells; include reuses tags.} Attach
    processes generators by ascending dimension to ensure boundary
    computability.

  \item \textbf{Smart extension.} The $\smartext$ algorithm infers
    boundary mappings via ogposet isomorphisms. This can fail if boundary
    shapes differ or if a tag maps inconsistently.

  \item \textbf{Totality is post-hoc.} Checked after all clauses,
    non-fatal.

  \item \textbf{Module idempotency.} A module is interpreted at most once.
    Import order can affect visible state.

  \item \textbf{Structural equality only.} Unit laws fail; middle-four
    interchange passes. Non-trivial equalities are higher cells.

  \item \textbf{Self-referencing identity.} A type's name inside its own
    complex resolves to the identity inclusion map.

  \item \textbf{Diamond imports.} Duplicate generators (same tag) from
    multiple imports are silently deduplicated.

  \item \textbf{Anonymous map convention.} In a named map definition
    $\kwlet\;f \mathbin{::} T = \ldots$, the annotation $T$ identifies
    the \emph{source}. In a diagram-level anonymous map
    $(\kwmap\;\ldots \mathbin{::} T)$, the annotation $T$ identifies
    where \emph{images} are evaluated---the opposite convention.

  \item \textbf{Type-level name uniqueness.} Name collisions are checked
    in complex bodies and local blocks but not in type blocks, where a
    duplicate $\kwlet$ silently overwrites.
\end{enumerate}

\end{document}
